# Terms and definitions {.unnumbered}

```{r}
#| label: setup
#| echo: false
#| output: false

library(here)
source(here::here("R/quarto-setup.R"))
```

```{r}
#| echo: false
#| output: asis

quarto_status("polishing")
```

* __Complex system__: There are several definitions. Here are some that I found to be of use:
	* "Systems that don't yield to compact forms of representation or description." [David Krakauer apud @mitchell2013]
	* "A system of many interacting parts where the system is more than just the sum of its parts" [Mark Newman apud @mitchell2013]
	* Systems with many connected agents that interact and exhibit self-organization and emergence behavior, all without the need for a central controller (adapted from Camilo Rodrigues Neto's definition, supervisor of this thesis).
	* Dialectics at its finest (my working definition).
* System theory: Two definitions can be of use:
	* Science or discipline that investigates models, principles, and laws that are valid to systems in general [@bertalanffy1968].
	* "The attempt of a reductionist scientific tradition to come to terms with complexity, nonlinearity, and change through sophisticated mathematical and computational techniques, _a groping toward a more dialectical understanding_ that is held back by its philosophical biases and the institutional and economic contexts of its development." [@levins1998]
